---
title: "Using tensoRphylo"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Using tensoRphylo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8} -->
---

```{r setup, echo = FALSE, message=FALSE}
library(tensoRphylo)
```

Welcome to the quickstart guide for the `R` package `tensoRphylo`, which is an `R` interface to the `C++` library `TensorPhylo` (note the difference in capitalization).
This library is designed to compute probabilities of trees (and discrete character data) under a general class of state-dependent sampled-birth-death models.
This general class of models includes effectively all state-dependent (and state-independent) sampled-birth-death models currently available in phylogenetics and phylogenetic epidemiology.

`TensorPhylo` is a not a _method_, and this package is not intended for directly fitting models to data.
Rather, `TensorPhylo` is a high-performance implementation of a _likelihood function_; it is up to you (or other package developers!) to implement the machinery to fit models of interest using this likelihood function.
The `tensoRphylo` package uses `Rcpp` to expose functionality from `TensorPhylo` to `R` users.
The purpose of this guide is to show `R` users how to use `tensoRphylo` in their own `R` code.
It is also possible to use the `TensorPhylo` library itself in `R` packages using `Rcpp`, but that is a topic for another guide.

# Reading in data

Before we start using `tensoRphylo`, we need a tree and some discrete character data.
For more information about file formats, see the [Data Formats](articles/dataformat.html) page.

We read in a simple tree in `ape::phylo` format provided in `inst/testdata`.
```{r fig.height = 7, fig.width = 10, out.width = "95%", fig.align = "center"}
# path to tree file
tree_file <- system.file("testdata", "extant_tree.nex", 
                         package = "tensoRphylo")

# read the tree
tree <- ape::read.nexus(tree_file)
```

Now we read in some discrete-character data.
```{r}
# path to data file
data_file <- system.file("testdata", "extant_data.nex", 
                         package = "tensoRphylo")

# read the data
data <- readNexusData(data_file)
```

# Creating a `tensoRphylo` instance

Now it's time to make a `tensoRphylo::TensorPhyloInstance`.
This `R` object mediates all interactions between `R` and the `TensorPhylo` backend library.
To make a new instance, we provide the tree and character data.
```{r}
# make a default tensorphylo instance
tp <- makeTensorPhylo(tree, data)
```
(`makeTensorPhylo` will do some automatic checks to make sure the tree and data are appropriately formatted.
See [Data Formats: Conformity](articles/dataformat.html#conformity) for more details.)

You can also skip the `data` argument and instead just provide a `nstates` argument.
This will create a "blank" data matrix with the `nstates` states (basically a hidden-character model with `nstates` hidden states).
```{r}
# make a default tensorphylo instance
tp <- makeTensorPhylo(tree, nstates = 2)
```

The instance begins as a state-independent constant-rate Yule model ($\lambda = 1$, $\mu = 0$, $\rho = 1$) for the tree, and a constant-rate _Mk_ model for the character data (transition rate $\eta = 0.1$).
We can already use the instance to compute a likelihood under this model.
```{r}
tp$computeLogLikelihood()
```
Of course, we will generally not be using the default model, let alone the default parameters! 
We'll see how to specify models and parameters next.

# Specifying models and parameters

Each model component (_e.g._, the speciation rates, extinction rates, transition rates, etc.) can be modified with specific models and parameter values.
In general, these processes may vary among character states and/or time, but we'll begin with the simple, constant-rate models.

## Constant-rate birth-death models

The constant-rate birth-death model assumes that species arise with rate $\lambda$, go extinct with rate $\mu$, and are sampled at the present with probability $\rho$.
To compute the likelihood of the tree under such a model, we'll create a `TensorPhyloInstance` without character data (but the minimum number of missing states):
```{r}
# our initial instance
tp <- makeTensorPhylo(tree, nstates = 2)
```

We can now set our desired parameter values like so:
```{r}
# set some parameter values
tp$setLambdaConstant(1.2)
tp$setMuConstant(0.5)
tp$setRhoPresent(0.5)

# compute the likelihood
tp$computeLogLikelihood()
```
Here, "constant" indicates that the rates are the same over time and among states; `setRhoPresent` specifies the sampling fraction at the present (since other models may include sampling at multiple time points).
(Because the character data are all missing and the diversification rates do not vary among discrete states, this is effectively a constant-rate birth-death model.)

To compute a likelihood under a new parameter value (or set of parameter values), don't recreate the instance.
Instead, just set the parameters as we did above and recompute the likelihood:
```{r}
# set different parameter values
tp$setLambdaConstant(0.9)
tp$setMuConstant(0.23)

# compute the likelihood
tp$computeLogLikelihood()
```

## Constant-rate birth-death models with character data

Now we'll see how to model discrete character evolution.
We begin by recreating the `TensorPhyloInstance` with the binary character data we loaded previously.
```{r}
# make the tensorphylo instance with character data
tp <- makeTensorPhylo(tree, data)
```

We'll again use a simple constant-rate birth-death model, but now we'll model the evolution of the discrete character as well.
As before, we'll set some basic birth-death model parameters.
```{r}
# set diversification parameters
tp$setLambdaConstant(0.5)
tp$setMuConstant(0.1)
tp$setRhoPresent(0.5)
```

We'll also specify a very simple model of character evolution by assuming transitions between states occur at the same rate.
In this case, the model has just one parameter, $\eta$.
```{r}
# set transition parameters
tp$setEtaConstantEqual(0.2)
```
Here, "constant" indicates that rates are the same over time, and "equal" indicates that the rates are equal among character states.

The likelihood is now the joint probability of the tree and character data, given the specified parameters:
```{r}
# compute the likelihood
tp$computeLogLikelihood()
```

Alternatively, we could allow rates of change to be different among character states.
We specify this type of model by first creating a `tensoRphylo::RateMatrix` object, which describes the instantaneous rate of change from state _i_ (in the row) and state _j_ (in the column).
```{r}
# create an empty rate matrix with the appropriate number of states
H <- makeRateMatrix(num_states = 2)

# specify the rate from state 1 to state 2
H[1,2] <- 0.1
H[2,1] <- 0.2

# print the matrix
H
```
Note that the rows of this matrix sum to zero; the diagonal elements are automatically computed when we set the off-diagonal rates.

We can then provide this rate matrix to the `TensorPhyloInstance`
```{r}
# set the rate matrix
tp$setEtaConstantUnequal(H)

# compute the likelihood
tp$computeLogLikelihood()
```
Here, `setEtaConstantUnequal` refers to the fact that rates of change are unequal among character states.

## State-dependent models

Now, we want to specify a more complex model: one that allows the diversification rates to vary among states.
Let's imagine we want to allow the speciation rate, $\lambda$, to vary among two binary states.
We can specify this model by creating a vector of speciation rates (one per state):
```{r}
# specify the state-dependent speciation rates
lambdas <- c(0.5, 1.0)

# set the parameters
tp$setLambdaStateDependent(lambdas)

# compute the likelihood
tp$computeLogLikelihood()
```
Here, `setLambdaStateDependent` means that the speciation rates depend on the character states.
**Note that we did not change the extinction rates or sampling fractions, so this model still assumes that they are the same for all character states.**

We can specify state-dependent extinction rates and sampling probabilities likewise:
```{r}
# specify the state-dependent extinction rates
mus <- c(0.05, 0.1)
tp$setMuStateDependent(mus)

# specify the state-dependent sampling probabilities
rhos <- c(0.3, 0.7)
tp$setRhoPresentStateDependent(rhos)

# compute the likelihood
tp$computeLogLikelihood()
```

## Time-dependent models

We can also let rates vary over time according to a discrete rate function.
Say we want to assume the speciation rate is the same among character states, but changes at a particular time point, $t$.
We can specify such a model like so:
```{r}
# specify the rate-change time and the rates
# NOTE: this undoes the state-dependent model we previously specified!
lambda_change_times <- c(1)
lambdas <- c(0.1, 0.5)

# set the parameters
tp$setLambdaTimeDependent(lambda_change_times, lambdas)

# compute the likelihood
tp$computeLogLikelihood()
```
In this case, time indicates how long ago the change occurred, measured from the present ($t = 0$).

There can be any number of rate-change times, but they must be ordered in increasing age.
There needs to be one more rate than rate change time; the first rate is the rate between the present and the first rate-change time, the second rate is the rate between the first rate-change time and the next, etc.
The last interval is open on the right (to the past).
For example, here is a model with three rates:
```{r}
# specify the rate-change time and the rates
lambda_change_times <- c(1, 4)
lambdas <- c(0.1, 0.5, 0.3)

# set the parameters
tp$setLambdaTimeDependent(lambda_change_times, lambdas)

# compute the likelihood
tp$computeLogLikelihood()
```

When multiple parameters are time-dependent, they do not need to share the same rate change times.
For example, here is a model where speciation and transition rates change at different times:
```{r}
# specify the speciation rates
lambda_change_times <- c(1, 4)
lambdas <- c(0.1, 0.5, 0.3)

# specify the transition rates
eta_change_times <- 5
etas <- c(0.1, 0.2)

# set the parameters
tp$setLambdaTimeDependent(lambda_change_times, lambdas)
tp$setEtaTimeDependentEqual(eta_change_times, etas)

# compute the likelihood
tp$computeLogLikelihood()
```

Transition rates can vary among states and over time, like so:
```{r}
# specify the transition-rate-change times
eta_change_times <- 5

# specify transition rates in the first interval
H_1 <- makeRateMatrix(num_states = 2)
H_1[1,2] <- 0.1
H_1[2,1] <- 0.2

# specify transition rates in the second interval
H_2 <- makeRateMatrix(num_states = 2)
H_2[1,2] <- 0.2
H_2[2,1] <- 0.1

# make the vector of rate matrices
Hs <- c(H_1, H_2)

# specify the transition rates
tp$setEtaTimeDependentUnequal(eta_change_times, Hs)

# compute the likelihood
tp$computeLogLikelihood()
```

## Time-and-state-dependent models

It's possible to specify time-and-state-dependent models in `tensoRphylo`, that is, models where diversification rates vary among states, but where the relationship between diversification rates and states itself varies over time.
This requires that we specify a _matrix_ of diversification rates, where the columns correspond to the rates per state, and the rows correspond to time intervals.

```{r}
# specify the speciation rates
lambda_change_times <- c(1, 4)
lambdas <- matrix(0, nrow = 3, ncol = 2) # three time intervals, two states
lambdas[1,] <- c(0.1, 0.2) # speciation rates in the first interval
lambdas[2,] <- c(0.2, 0.1) # speciation rates in the second interval
lambdas[3,] <- c(0.4, 0.6) # speciation rates in the third interval

# set the parameters
tp$setLambdaTimeStateDependent(lambda_change_times, lambdas)

# compute the likelihood
tp$computeLogLikelihood()
```

## Cladogenetic models

`tensoRphylo` implements models where states may change at speciation events, such that one or both of the descendant lineages can have a different state than their ancestor, _i.e._, where state changes are _cladogenetic_.
This is a common feature of biogeographic and chromosomal models.

## Models with mass events

# Other settings

## Error handling

<!-- By default, the `TensorPhyloInstance` will complain if you try to provide inappropriate parameter values, _e.g._, rate parameters less than zero, or probability parameters greater than 1 or less than zero: -->
<!-- ```{r error=TRUE} -->
<!-- tp$setLambdaConstant(-1.2) -->
<!-- tp$setRhoPresent(1.5) -->
<!-- ``` -->
<!-- (You can turn this behavior off if you want to do error handling on your own with `tp$setSafeMode(FALSE)`.) -->











<!-- We can now write a function of the parameters $\lambda$ and $\mu$, and estimate them by maximum likelihood. -->
<!-- ```{r} -->
<!-- constantRateLikelihood <- function(pars) { -->

<!--   # get the parameters -->
<!--   lambda <- pars[1] -->
<!--   mu     <- pars[2] -->

<!--   # make sure that parameters are not negative -->
<!--   if ( lambda < 0 || mu < 0 ) { -->
<!--     return(Inf) -->
<!--   } -->

<!--   # set the parameters -->
<!--   tp$setLambdaConstant(lambda) -->
<!--   tp$setMuConstant(mu) -->

<!--   # compute the likelihood -->
<!--   -tp$computeLogLikelihood() -->

<!-- } -->

<!-- # estimate the parameters by maximum likelihood -->
<!-- fit <- optim(c(1,0.5), constantRateLikelihood) -->

<!-- # report the MLEs -->
<!-- fit$par -->
<!-- ``` -->
