---
title: "API documentation"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: ../references.bib
vignette: >
  %\VignetteIndexEntry{API documentation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8} -->
---

<!-- This stuff is just for me for troubleshooting inside this document. -->

```{r setup, echo = FALSE, message=FALSE}
library(tensoRphylo)
```

```{r, echo = FALSE}
# read the tree
tree <- ape::read.nexus(system.file("testdata", "extant_tree.nex", package = "tensoRphylo"))

# read character data
data <- readNexusData(system.file("testdata", "extant_data.nex", package = "tensoRphylo"))

# make the instance
tp <- makeTensorPhylo(tree, data)
```

# 1 Notation and data types

`tensoRphylo` is very particular about the types and sizes of arguments you provide.
Throughout this documentation I will use _data types_ as proxies for variables to make it clear what `tensoRphylo` expects.
I will indicate the data type and the package of origin as `package::type`.
I will also do this for return types (when assigning a result to an object) to make it clear what data types are generated by `tensoRphylo` objects.

When relevant, I will also provide an indication of the correct dimensionality of the object using parantheses.
The parantheses will contain comma-separate values that indicate the size of each dimension of the object.
For example, when I mean to indicate that an argument should be a single numeric value (a scalar), I will write `base::numeric(1)`.
When I mean that the argument should be a numeric vector, I will write `base::numeric(s)` to indicate a vector of length `s`.
When I mean that the argument should be a numeric matrix, I will write `base::matrix(r, c)`, where `n` indicates the number of rows and `c` indicates the number of columns.

The dimensions of many of the objects are determined by the number of tips in the tree or the number of states in the data matrix.
I will use `num_tips` as a proxy for the number of tips, and `num_states` as a proxy for the number of states.

# 2 Constructors

Create a `TensorPhyloInstance` with the `makeTensorPhylo`.
You _must_ provide a phylogeny, and also either: 1) a character dataset, or 2) the number of missing character states.
See the [Input](dataformat.hmtl) for more details about data format.

```{r, eval = FALSE}
# create an instance with a tree and a character vector.
# supports ambiguity and missing data.
tp <- makeTensorPhylo(tree = ape::phylo,
                      data = base::character(num_tips))
```

```{r, eval = FALSE}
# create an instance with a tree and a numeric vector.
# does not support ambiguity or missing data.
tp <- makeTensorPhylo(tree = ape::phylo,
                      data = base::numeric(num_tips))
```

```{r, eval = FALSE}
# create an instance with a tree and a numeric matrix.
# supports ambiguity and missing data.
tp <- makeTensorPhylo(tree = ape::phylo,
                      data = base::matrix(num_tips, num_states))
```

```{r, eval = FALSE}
# create an instance with a tree but no data.
# create a missing character with `num_states` states.
tp <- makeTensorPhylo(tree       = ape::phylo,
                      num_states = base::numeric(1))
```

# 3 Computing probabilities

The core functionality of a `TensorPhyloInstance` is to compute the likelihood of the currently active model/parameters.
This includes the parameters of the diversification model, state-change model, and root frequency model, as well as the chosen conditional probability type, each of which is described below.
Compute the likelihood like so:
```{r, eval = FALSE}
# compute the log likelihood
base::numeric(1) <- tp$computeLogLikelihood()
```

## 3.1 Labeled trees

By default, a `TensorPhyloInstance` will compute the probability density of the _labeled_ tree and states, which includes a combinatorial constant, as described in @stadler2013can.
However, most comparable `R` implementations of state-dependent birth-death models (_e.g._ `diversitree` and `hisse`) do not include this constant.
To suppress this constant, do:
```{r, eval = FALSE}
# turn off the combinatorial constant
tp$setApplyTreeLikCorrection(FALSE)

# turn on the combinatorial constant (default)
tp$setApplyTreeLikCorrection(TRUE)
```



# 4 Lineage-specific diversification models

Lineage-specific diversification events occur independently among lineages, but can depend on the character state and/or time.
There are for types of events: speciation, extinction, sampling, and destructive sampling (simultaneous sampling and extinction).
Each type of event can occur according to one of four models: constant, state-dependent, time-dependent or time-and-state-dependent.
The type of model can be different for different types of events.

## 4.1 Time-dependent diversification models

The time-dependent models assume that rates are the same between discrete rate-change events.
The times of the rate-change events are measured backward in time, so that $t = 0$ is the present and $t > 0$ is some time in the past.
The times should be in ascending order, such that more recent events appear first.

A model with `n` rate-change times will have `n + 1` time intervals.
The last interval is "open" in the sense that it covers all time after the `n`<sup>th</sup> time point.
You will need to provide `n + 1` parameters (or sets of parameters), one per interval.
These should be ordered so that the first set of parameters corresponds to the first interval, the second corresponds to the second interval, etc.

## 4.2 Speciation events

During a speciation event, the affected lineage speciates, producing two new lineages.
If the model includes cladogenetic character change (see below), the lineage experiences cladogenetic evolution according to the cladogenetic process at that time; otherwise, the new lineages inherit the ancestor's state.
Speciation events are described by the parameter $\lambda$; the corresponding functions include `lambda` in their name.

### 4.2.1 Constant speciation events

```{r, eval = FALSE}
# specify a single speciation-rate parameter that applies to all
# states and all times
speciation_rate <- base::numeric(1)
tp$setLambdaConstant(speciation_rate)
```

### 4.2.2 State-dependent speciation events

```{r, eval = FALSE}
# specify a state-dependent speciation-rate parameter that does
# not vary over time
speciation_rates <- base::numeric(num_states)
tp$setLambdaStateDependent(speciation_rates)
```

### 4.2.3 Time-dependent speciation events

```{r, eval = FALSE}
# specify a time-dependent speciation-rate parameter that does not 
# depend on the state
speciation_change_times <- base::numeric(num_rate_shifts)
speciation_rates        <- base::numeric(num_rate_shifts + 1)
tp$setLambdaStateDependent(speciation_change_times, speciation_rates)
```

### 4.2.4 Time-and-state-dependent speciation events

```{r, eval = FALSE}
# specify a time-dependent speciation-rate that depends 
# on the state
speciation_change_times <- base::numeric(num_rate_shifts)
speciation_rates        <- base::matrix(num_rate_shifts + 1, num_states)
tp$setLambdaTimeStateDependent(speciation_change_times, speciation_rates)
```

## 4.3 Extinction events

During an extinction event, the affected lineages goes extinct.
Extinction events are described by the parameter $\mu$; the corresponding functions include `mu` in their name.

### 4.3.1 Constant extinction events

```{r, eval = FALSE}
# specify a single extinction-rate parameter that applies to all
# states and all times
extinction_rate <- base::numeric(1)
tp$setMuConstant(extinction_rate)
```

### 4.3.2 State-dependent extinction events

```{r, eval = FALSE}
# specify a state-dependent extinction-rate parameter that does
# not vary over time
extinction_rates <- base::numeric(num_states)
tp$setMuStateDependent(extinction_rates)
```

### 4.3.3 Time-dependent extinction events

```{r, eval = FALSE}
# specify a time-dependent extinction-rate parameter that does not 
# depend on the state
extinction_change_times <- base::numeric(num_rate_shifts)
extinction_rates        <- base::numeric(num_rate_shifts + 1)
tp$setMuStateDependent(extinction_change_times, extinction_rates)
```

### 4.3.4 Time-and-state-dependent extinction events

```{r, eval = FALSE}
# specify a time-dependent extinction-rate parameter that depends 
# on the state
extinction_change_times <- base::numeric(num_rate_shifts)
extinction_rates        <- base::matrix(num_rate_shifts + 1, num_states)
tp$setMuTimeStateDependent(extinction_change_times, extinction_rates)
```

## 4.4 Sampling events

During a sampling event, the affected lineages produces a sample.
Sampling events are described by the parameter $\phi$; the corresponding functions include `phi` in their name.

### 4.4.1 Constant sampling events

```{r, eval = FALSE}
# specify a single sampling-rate parameter that applies to all
# states and all times
sampling_rate <- base::numeric(1)
tp$setPhiConstant(sampling_rate)
```

### 4.4.2 State-dependent sampling events

```{r, eval = FALSE}
# specify a state-dependent sampling-rate parameter that does
# not vary over time
sampling_rates <- base::numeric(num_states)
tp$setPhiStateDependent(sampling_rates)
```

### 4.4.3 Time-dependent sampling events

```{r, eval = FALSE}
# specify a time-dependent sampling-rate parameter that does not 
# depend on the state
sampling_change_times <- base::numeric(num_rate_shifts)
sampling_rates        <- base::numeric(num_rate_shifts + 1)
tp$setPhiStateDependent(sampling_change_times, sampling_rates)
```

### 4.4.4 Time-and-state-dependent sampling events

```{r, eval = FALSE}
# specify a time-dependent sampling-rate parameter that that depends 
# on the state
sampling_change_times <- base::numeric(num_rate_shifts)
sampling_rates        <- base::matrix(num_rate_shifts + 1, num_states)
tp$setPhiTimeStateDependent(sampling_change_times, sampling_rates)
```

## 4.5 Destructive-sampling events

During a destructive-sampling event, the affected lineages produces a sample and simultaneously goes extinct.
Destructive-sampling events are described by the parameter $\phi$; the corresponding functions include `phi` in their name.

### 4.5.1 Constant destructive-sampling events

```{r, eval = FALSE}
# specify a single destructive-sampling-rate parameter that applies to all
# states and all times
destructive_sampling_rate <- base::numeric(1)
tp$setDeltaConstant(destructive_sampling_rate)
```

### 4.5.2 State-dependent destructive-sampling events

```{r, eval = FALSE}
# specify a state-dependent destructive-sampling-rate parameter that does
# not vary over time
destructive_sampling_rates <- base::numeric(num_states)
tp$setDeltaStateDependent(destructive_sampling_rates)
```

### 4.5.3 Time-dependent destructive-sampling events

```{r, eval = FALSE}
# specify a time-dependent destructive-sampling-rate parameter that does not 
# depend on the state
destructive_sampling_change_times <- base::numeric(num_rate_shifts)
destructive_sampling_rates        <- base::numeric(num_rate_shifts + 1)
tp$setDeltaStateDependent(destructive_sampling_change_times, 
                          destructive_sampling_rates)
```

### 4.5.4 Time-and-state-dependent destructive-sampling events

```{r, eval = FALSE}
# specify a time-dependent destructive-sampling-rate parameter that depends 
# on the state
destructive_sampling_change_times <- base::numeric(num_rate_shifts)
destructive_sampling_rates        <- base::matrix(num_rate_shifts + 1, num_states)
tp$setDeltaTimeStateDependent(destructive_sampling_change_times, 
                              destructive_sampling_rates)
```

# 5 Mass-diversification models

In contrast to lineage-specific-diversification events, mass-diversification events affect all lineages living at a given time.
There are four types of mass-diversification events: mass-speciation, mass-extinction, mass-sampling and mass-destructive-sampling.
Each event is associated with a magnitude, which is the probability that the event occurs to a given lineages.
These magnitude may be independent of the state of the lineage (a "constant" event), or it may be state-dependent.

Each mass-diversification event occurs at a particular time with a particular magnitude (or set of state-dependent magnitudes).
There can be multiple events of a given type, each with different event times.
The times of the events are measured backward in time, so that $t = 0$ is the present and $t > 0$ is some time in the past.
You should provide the times of the events in ascending order, such that more recent events appear first; the magnitudes should be ordered accordingly.

## 5.1 Mass-speciation events

Mass-speciation events are described by the parameter $\Upsilon$; the corresponding functions include `Upsilon` in their name.
During a mass-speciation event, all lineages in state $i$ alive at the time experience a speciation event with probability $\Upsilon_i$.
If the model includes cladogenetic character change (see below), the speciating lineages experience cladogenetic evolution according to the cladogenetic process at that time; otherwise, both daughters inherit the ancestor's state.

### 5.1.1 State-independent mass-speciation events

```{r, eval = FALSE}
# specify state-independent mass-speciation events at given times
mass_speciation_times       <- base::numeric(num_events)
mass_speciation_probability <- base::numeric(num_events)
tp$setUpsilonConstant(mass_speciation_times, mass_speciation_probability)
```

### 5.1.2 State-dependent mass-speciation events

```{r, eval = FALSE}
# specify state-dependent mass-speciation events at given times
mass_speciation_times       <- base::numeric(num_events)
mass_speciation_probability <- base::matrix(num_events, num_states)
tp$setUpsilonStateDependent(mass_speciation_times, mass_speciation_probability)
```

## 5.2 Mass-extinction events

Mass-extinction events are described by the parameter $\Gamma$; the corresponding functions include `Gamma` in their name.
During a mass-extinction event, all lineages in state $i$ alive at the time experience an extinction event with probability $\Gamma_i$.

### 5.2.1 State-independent mass-extinction events

```{r, eval = FALSE}
# specify state-independent mass-extinction events at given times
mass_extinction_times       <- base::numeric(num_events)
mass_extinction_probability <- base::numeric(num_events)
tp$setGammaConstant(mass_extinction_times, mass_extinction_probability)
```

### 5.2.2 State-dependent mass-extinction events

```{r, eval = FALSE}
# specify state-dependent mass-extinction events at given times
mass_extinction_times       <- base::numeric(num_events)
mass_extinction_probability <- base::matrix(num_events, num_states)
tp$setGammaStateDependent(mass_extinction_times, mass_extinction_probability)
```

## 5.3 Mass-sampling events

Mass-sampling events are described by the parameter $R$; the corresponding functions include `Rho` in their name.
During a mass-sampling event, all lineages in state $i$ alive at the time experience a sampling event with probability $R_i$.

### 5.3.1 State-independent mass-sampling at the present

```{r, eval = FALSE}
# specify a state-independent mass-sampling event at the present (t = 0)
mass_sampling_probability <- base::numeric(1)
tp$setRhoPresent(mass_sampling_probability)
```

### 5.3.2 State-dependent mass-sampling at the present

```{r, eval = FALSE}
# specify a state-dependent mass-sampling event at the present (t = 0)
mass_sampling_probability <- base::numeric(num_states)
tp$setRhoPresentStateDependent(mass_sampling_probability)
```

### 5.3.3 State-independent mass-sampling events

```{r, eval = FALSE}
# specify state-independent mass-sampling events at given times
mass_sampling_times       <- base::numeric(num_events)
mass_sampling_probability <- base::numeric(num_events)
tp$setRhoConstant(mass_sampling_times, mass_sampling_probability)
```

### 5.3.4 State-dependent mass-ampling events

```{r, eval = FALSE}
# specify state-dependent mass-sampling events at given times
mass_sampling_times       <- base::numeric(num_events)
mass_sampling_probability <- base::matrix(num_events, num_states)
tp$setRhoStateDependent(mass_sampling_times, mass_sampling_probability)
```

## 5.4 Mass-destructive-sampling events

Mass-destructive-sampling events are described by the parameter $\Xi$; the corresponding functions include `Xi` in their name.
During a mass-destructive-sampling event, all lineages in state $i$ alive at the time experience a destructive-sampling event with probability $\Xi_i$.

### 5.4.1 State-independent mass-destructive-sampling events

```{r, eval = FALSE}
# specify state-independent mass-destructive-sampling events at given times
mass_destructive_sampling_times       <- base::numeric(num_events)
mass_destructive_sampling_probability <- base::numeric(num_events)
tp$setXiConstant(mass_destructive_sampling_times, 
                 mass_destructive_sampling_probability)
```

### 5.4.2 State-dependent mass-destructive-sampling events

```{r, eval = FALSE}
# specify state-dependent mass-sampling events at given times
mass_destructive_sampling_times       <- base::numeric(num_events)
mass_destructive_sampling_probability <- base::matrix(num_events, num_states)
tp$setXiStateDependent(mass_destructive_sampling_times, 
                       mass_destructive_sampling_probability)
```

# 6 State-change models

There are two types of state-change events: "anagenetic" events, which occur along lineages and are not associated with speciation events, and "cladogenetic" events, which are associated with speciation events.
Both types of events can occur according to two models: constant and time-dependent.

## 6.1 Time-dependent state-change models

The time-dependent models assume that state-change events are the same between discrete shift events.
The times of the shift events are measured backward in time, so that $t = 0$ is the present and $t > 0$ is some time in the past.
The times should be in ascending order, such that more recent events appear first.

A model with `n` shift events will have `n + 1` time intervals.
The last interval is "open" in the sense that it covers all time after the `n`<sup>th</sup> time point.
You will need to provide `n + 1` parameters (or sets of parameters), one per interval.
These should be ordered so that the first set of parameters corresponds to the first interval, the second corresponds to the second interval, etc.

## 6.2 Anagenetic events

During an anagenetic state-change event, a lineage changes from state _i_ to state _j_.
The rate at which these events occurs is described by a rate matrix, $H$ (denoted `Eta` in the functions below).
The rate at which a lineage in state _i_ changes to state _j_ is $\eta_{ij}$.
This rate matrix is analogous to the instantaneous-rate matrix, $Q$, used in standard continuous-time Markov models used to describe discrete-character evolution.
See the documentation for `tensoRphylo::RateMatrix` for more information about rate matrices.

### 6.2.1 Constant, symmetrical anagenetic events

```{r, eval = FALSE}
# specify constant-rate, symmetrical anagenetic process
transition_rate <- base::numeric(1)
tp$setEtaConstantEqual(transition_rate)
```

### 6.2.2 Constant, asymmetrical anagenetic events

```{r, eval = FALSE}
# specify constant-rate, asymmetrical anagenetic process
transition_matrix <- tensoRphylo::RateMatrix(num_states, num_states)
tp$setEtaConstantUnequal(transition_matrix)
```

### 6.2.3 Time-dependent, symmetrical anagenetic events

```{r, eval = FALSE}
# specify time-dependent, symmetrical anagenetic process
transition_change_times <- base::numeric(num_rate_shifts)
transition_rates        <- base::numeric(num_rate_shifts + 1)
tp$setEtaTimeDependentEqual(transition_change_times, transition_rates)
```

### 6.2.4 Time-dependent, asymmetrical anagenetic events

```{r, eval = FALSE}
# specify time-dependent, symmetrical anagenetic process
transition_change_times <- base::numeric(num_rate_shifts)
transition_matrix       <- tensoRphylo::RateMatrix(num_states, num_states)
transition_matrices     <- tensoRphylo::RateMatrixList(num_rate_shifts + 1)
tp$setEtaTimeDependentUnequal(transition_change_times, transition_matrices)
```

### 6.2.5 Mass-anagenetic events

A mass-anagenetic event occurs at a particular time, $t$; all lineages alive at time $t$ experience anagenetic change according to the probability matrix $Z$; $\zeta_{ij}$ is the probability that a lineage in state _i_ changes to state _j_ during the event.
See the documentation for `tensoRphylo::ProbabilityMatrix` for more information about probability matrices (and lists of probability matrices).

```{r, eval = FALSE}
# specify mass-anagenetic events
mass_transition_times    <- base::numeric(num_transition_events)
mass_transition_matrix   <- tensoRphylo::ProbabilityMatrix(num_states, num_states)
mass_transition_matrices <- tensoRphylo::ProbabilityMatrixList(num_transition_events)
tp$setZeta(mass_transition_times, mass_transition_matrices)
```

### 6.2.6 Mass-extinction-and-anagensis events

We can model simultaneous mass-extinction and anagenetic character change.
To do so, we simply combine the mass-extinction and mass-anagenesis parameters, $\Gamma$ and $Z$.
You must provide one $Z$ matrix per mass-extinction event.
NOTE: We originally implemented mass-anagenetic events to accommodate changes in state associated with mass-extinction events.
Consequently, you cannot model mass-anagenetic events and mass-extinction-and-anagenesis events separately.
We intend to relax this dependence in the future.

In these examples, state-independent and state-dependent refers to the magnitudes of the mass-extinction events.

#### 6.2.6.1 State-independent mass-extinction event with anagenesis

```{r, eval = FALSE}
# specify state-independent mass-extinction-and-anagensis events at given times
mass_extinction_times       <- base::numeric(num_events)
mass_extinction_probability <- base::numeric(num_events)
mass_transition_matrices    <- tensoRphylo::ProbabilityMatrixList(num_events)
tp$setGammaAndZetaConstant(mass_extinction_times, 
                           mass_extinction_probability,
                           mass_transition_matrices)
```

#### 6.2.6.2 State-dependent mass-extinction event with anagenesis

```{r, eval = FALSE}
# specify state-dependent mass-extinction-and-anagensis events at given times
mass_extinction_times       <- base::numeric(num_events)
mass_extinction_probability <- base::matrix(num_events, num_states)
mass_transition_matrices    <- tensoRphylo::ProbabilityMatrixList(num_events)
tp$setGammaAndZetaStateDependent(mass_extinction_times, 
                                 mass_extinction_probability,
                                 mass_transition_matrices)
```

## 6.3 Cladogenetic events

During a cladogenetic state-change event, a lineage in state _i_ leaves one lineage in state _j_ and another lineage in state _k_.
In `tensoRphylo`, cladogenetic rates do not occur with a _rate_; rather, they occur with some probability whenever a speciation event occurs.
The probability that a lineage in state _i_ leaves one lineage in state _j_ and another in state _k_ is $\omega_{ijk}$.
We represent these probabilities in a three-dimensional array, $\Omega$ (denoted `Omega` below), where the first dimension corresponds to the ancestral state, the second dimension to the state of one daughter lineage, and the third dimension to the state of the second daughter lineage.
See the documentation for `tensoRphylo::CladoEvents` for more information about creating these arrays.

(Other software typically combines the speciation rate with the cladogenetic event probability.
For example, in `diversitree`, the rate at which a lineage in state _i_ speciates and produces one lineage in state _j_ and another in state _k_ is denoted $\lambda_{ijk}$.
The equivalent value in `tensoRphylo` is $\lambda_i \times \omega_{ijk}$.
We chose this alternative parameterization in `TensorPhylo` because it makes it more concise to express a non-cladogenetic model as a special case of a cladogenetic model.)

### 6.3.1 Constant cladogenetic events

```{r, eval = FALSE}
# specify constant-rate cladogenetic process
cladogenetic_array <- tensoRphylo::CladoEvents(num_states, 
                                               num_states, 
                                               num_states)
tp$setOmegaConstant(cladogenetic_array)
```

### 6.3.2 Time-dependent cladogenetic events

```{r, eval = FALSE}
# specify time-dependent cladogenetic process
cladogenetic_change_times <- base::numeric(num_rate_shifts)
cladogenetic_arrays <- tensoRphylo::CladoEventsList(num_rate_shifts + 1)
tp$setOmegaTimeDependent(cladogenetic_change_times, cladogenetic_arrays)
```

### 6.3.3 Mass-cladogenetic events

Because cladogenetic state-change events only occur at speciation events, mass-cladogenetic state-change events are not explicitly modeled.
Rather, they are _implied_ if you use a model with mass-speciation events and cladogenetic state changes.
When a lineage speciates during a mass-speciation event at time $t$, that lineage experiences cladogenetic state changes just as it would during a normal speciation event at that time (including any time-dependence in the cladogenetic state-change process).

# 7 Root prior models

The root prior parameter controls the probability that the tree begins in each possible state.
By default, we assume a flat root frequency prior.
However, you can also use a specified root frequency (_i.e._, something other than a uniform), or the quasistationary frequency.

## 7.1 Uniform root prior (default)

```{r, eval = FALSE}
# specify uniform probability over all states at the beginning of the tree
tp$setRootPriorFlat()
```

## 7.2 Non-uniform root prior

```{r, eval = FALSE}
# specify probability distribution over all states at the beginning of the tree
tp$setRootPrior(base::numeric(num_states))
```

## 7.3 Quasistationary frequency prior

The quasistationary frequency is the frequency of lineages in each possible state after a long (infinite) period of time, taking into account differential diversification rates and the assumed state-change model (including both anagenesis and cladogenesis).

The quasistationary frequency assumes the tree evolves under a given process for an infinite amount of time (assuming it doesn't go extinct) but doesn't experience mass events of any type.
If you use this option, `TensorPhylo` will use the parameters at the beginning of the tree (the root or stem) to determine the quasistationary frequency.
Consequently, this option doesn't really make sense for models where the parameters changes over time, or when there are mass events of any type, because the process may not be quasi-stationary.

```{r, eval = FALSE}
# tell tensorphylo to use the quasistationary frequency for the root prior
tp$setQuasistationaryFrequencyMode(TRUE)
```

You can also query the `TensorPhyloInstance` for its quasistationary frequency for the parameters at a particular time, $t$.
```{r, eval = FALSE}
# check the quasistationary frequency of the parameters at a particular time
base::numeric(num_states) <- tp$getQuasiStationaryFrequency(base::numeric(1))
```

# 8 Conditional probabilities

Conditional probabilities represent ascertainment corrections induced by not studying some fraction of outcomes under the model.
For example, _conditioning on survival_ implies that we would not study trees that did not survive to the present.
The type of conditioning to use depends on whether the tree begins with 1 or 2 lineages, and how you choose to model sampling.
These options are discussed in more detail in the `tensoRphylo::conditionalProbability` documentation.

Conditional probability settings are implemented as named elements of the global `tensoRphylo::conditionalProbability` object.
To apply a particular correction, provide it to the `TensorPhyloInstance` like so:
```{r}
# condition the process on the two descendants of the root producing at least
# one extant descendant each. this is equivalent to conditioning on "survival"
# in most other software.
tp$setConditionalProbabilityType(conditionalProbability$ROOT_MRCA)
```

# 9 Stochastic mapping

## 9.1 Generate a sample of stochastic maps 

Simulate stochastic maps in `phytools` format.
This simulation can take a bit of time, but you can stop it early, returning the replicates that have completed.

```{r, eval = FALSE}
# simulate `reps` stochastic maps
phytools::multiSimmap(reps) <- tp$drawStochasticMaps(reps = base::numeric(1))
```

## 9.2 Simulate the average diversification rates per branch

This function computes the average diversification rate for each branch using stochastic mapping.
For each stochastic map, computes the diversification rate on each branch as the weighted average of the state-dependent diversification- and sampling-rate parameters.
Returns a list of matrices, each of which corresponds to the branch-specific (in rows) average diversification and sampling rates (columns).

```{r, eval = FALSE}
# simulate `reps` branch rates
base::list<base::matrix(num_branches,5)>(reps) <- tp$drawBranchRates(reps = base::numeric(1))
```

# 10 Other settings

## 10.1 Parallel processing

`TensorPhylo` implements parallel processing with `OpenMP`.
To specify the number of threads to use, simply do as follows:
```{r, eval = FALSE}
# specify the number of threads to use
tp$setNumberOfThreads(base::numeric(1))
```

Note that only certain likelihood algorithms will actually use multiple threads; see the [Likelihood algorithms](#approximators) section for details.
Depending on the size of the tree and the number of character states, parallel processing will sometimes not be worthwhile, or even be detrimental.
Parallel processing works the best with large numbers of states and/or with cladogenetic models.
See the [Multiprocessing performance comparison](comparison.html#multicore) for some idea of what to expect.

## 10.2 Numerical integrators { #integrators }

`TensorPhylo` uses `boost/odeint` numerical integrators for solving ODEs.
These integrators vary in terms of their speed and accuracy, but the default one is generally quite good.
For more information, see the `tensoRphylo::integrationScheme` documentation and the [`boost/odeint`](https://www.boost.org/doc/libs/1_66_0/libs/numeric/odeint/doc/html/index.html) documentation.
See our [Integrator comparison](comparison.html#integrator_comparison) for an idea about how these integrators compare.

Numerical integration options are implemented as named elements of the global `tensoRphylo::integrationScheme` object.
All of these integrators are adaptive, so you are not required to specify a certain `dt`.
To use a particular integrator, provide it to the `TensorPhyloInstance` like so:
```{r, eval = FALSE}
# use the Runge-Kutta 54 (AKA Runge-Kutta-Fehlberg) adaptive
# numerical integrator (also the default, and equivalent to the
# scheme used by diversitree)
tp$setIntegrationScheme(integrationScheme$RUNGE_KUTTA54)
```

## 10.3 Likelihood algorithms { #approximators }

`TensorPhylo` implements a few different algorithms for computing the likelihood.
These algorithms differ in how they share information and/or parallelize calculations across branches.
The algorithm options are implemented as named elements of the global `tensoRphylo::approximatorVersion` object.
The default setting, `AUTO_TUNING`, initially decides which algorithm to use based on performance modeling calibrated by extensive simulation; it then dynamically chooses the best-performing algorithm (and number of cores, if you're allowing more than 1) for a particular dataset over the course of many likelihood calculations.
For this reason, we recommend always using this algorithm

To use a particular algorithm, provide it to the `TensorPhyloInstance` like so:
```{r, eval = FALSE}
# use the SEQUENTIAL_BRANCHWISE algorithm (similar to the one used by
# diversitree)
tp$setLikelihoodApproximator(approximatorVersion$SEQUENTIAL_BRANCHWISE)
```

## 10.4 Error handling

By default, `tensoRphylo` checks the parameters you provide to make sure they conform to certain requirements; for example, that rate parameters are positive, that the number of rate parameters matches the number of states and/or rate changes, etc.
You can toggle this behavior (for example, if you want to implement your own error handling) like so:
```{r, eval = FALSE}
# turn off parameter checking
tp$setSafeMode(FALSE)

# or turn on parameter checking
tp$setSafeMode(TRUE)
```

## 10.5 Debugging

You can use built-in `TensorPhylo` interactive debug mode to have the internal `TensorPhylo` object report what parameter values, settings, and data are being passed from an `R` session.

To report to the screen, use:
```{r, eval = FALSE}
# report data/parameters/settings to the screen as they are provided
tp$setDebugMode(1)
```

To report to a file, use:
```{r, eval = FALSE}
# report data/parameters/settings to a file as they are provided
tp$setDebugFile(file = base::character(1))
```

To turn off debug mode, use:
```{r, eval = FALSE}
# turn off debug mode
tp$setDebugMode(0)
```

# References {-}


