---
title: "Input formats"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Input formats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8} -->
---

```{r setup, echo = FALSE, message = FALSE}
library(tensoRphylo)
```

The input formats used by `tensorPhylo` are fairly standard and intended to require minimal preprocessing: everything we do is native to standard Newick, `NEXUS`, and `ape::phylo` formats.

# Tree

The tree **must** be a rooted tree in `ape::phylo` format.
In general, `ape::read.nexus` does an excellent job of parsing valid Newick strings with the below features such that they can be used in `tensoRphylo` without any additional processing.

Trees **may** have any combination of extant and/or extinct tips, as well as sampled ancestors. (In `ape::phylo`, sampled ancestors are represented as nodes with a single descendant.)

The tree **may** have a stem branch subtending the root, in which case `tensoRphylo` will assume the process begins with one lineage at the start of this stem branch.
Otherwise, `tensoRphylo` will assume the process begin with the two lineages at the root of the tree.
The length of the stem branch is stored in the `root.edge` element of an `ape::phylo` object.

## A closer look at a tree with extinct tips and sampled ancestors

The sampled-ancestor tree (with extant and/or extinct samples and samples along branches) is the most complex type of tree produced by the general modeling framework implemented in `tensoRphylo`; all other trees (_e.g._, trees of extant taxa only) are a special case of this tree.
You can find a nexus file with a valid Newick representation of a sampled-ancestor tree in the package data ([link to raw file](https://github.com/mikeryanmay/tensoRphylo/blob/main/inst/testdata/sampled_ancestor_tree.nex)):
```{r}
# path to tree file
tree_file <- system.file("testdata", "sampled_ancestor_tree.nex",  
                         package = "tensoRphylo")

# read the tree
tree <- ape::read.nexus(tree_file)

# print the tree
tree
```
Notice that this tree has more internal nodes than tips, because some of the nodes (the sampled ancestors) have only one descendant.

Inspecting the node labels, we see that some nodes are labeled `""` and others are labeled with strings:
```{r}
# print node labels
tree$node.label
```
The nodes with strings are the sampled ancestors; these node labels are used to associate the sampled ancestor with character data (discussed in the next section).

This tree also has a stem:
```{r}
# print the root edge
tree$root.edge
```

We can plot the tree to see that there are sampled ancestors.
```{r, fig.height = 7, fig.width = 10, out.width = "95%", fig.align = "center", fig.cap = "A tree with extant and extinct tips, as well as sampled ancestors (dots along branches)."}
# plot the tree with a root edge
ape::plot.phylo(tree, no.margin = TRUE, root.edge = TRUE, underscore = TRUE)

# plot the sampled ancestors with names
sampled_ancestors <- (1:tree$Nnode + Ntip(tree))[tree$node.label != ""]
ape::nodelabels(node = sampled_ancestors, text = tree$node.label[tree$node.label != ""], adj = c(0, -1), frame = "none", srt = 35)
ape::nodelabels(node = sampled_ancestors, pch  = 19)
```

# Character data

The discrete character data can either be in the "traditional" format, _i.e._, a named vector specifying the state for each sample, or a numeric matrix with rows corresponding to the samples and the columns correspond to the states.
I briefly describe each of these formats below.

## Named vector

This format should be familiar to people who have used `ape`, `phytools`, `diversitree`, etc.
In this format, the data are represented as a vector of numeric, integer, or string variables, each element named by the corresponding sample in the tree.
While it is often convenient, the main drawback of this format is that `tensoRphylo` must figure out the state space of the model from the data.
As a consequence, `tensoRphylo` will only include states that are present in the data.
This will be inappropriate when you know there are states that are not present in any sample; for example, if the state is the number of chromosomes, then logically the state space should include all integers, even if not every number of chromosomes is observed in the data.
I provide an example of a dataset with this format in the package data:
```{r}
# path to character data file
data_file <- system.file("testdata", "sampled_ancestor_data_vec.Rda",  
                         package = "tensoRphylo")

# read the character data
data <- readRDS(data_file)

data
```

`tensoRphylo` has some **added functionality** when provided with a **named vector of strings**.
First, it accommodates _missing_ data, represented as a `?` or `-`.
Second, it accommodates _ambiguous_ data, _i.e._, when a sample could be one of many possible (perhaps hidden) states.
Ambiguous data is represented as a `/`-delimited string of the possible states, as you see above.
We can use the internal function `tensoRphylo:::.char.vector.to.table` to see how `tensoRphylo` translates this type of data into the numeric matrix format, described next:
```{r}
# make some missing data
data[2] <- "?"
data[3] <- "-"

# convert to numeric matrix and print the result
tensoRphylo:::.char.vector.to.table(data)
```
Note that the top three rows (corresponding to the samples we set to ambiguous/missing data) are now all `1`, because both states are possible.

## Numeric matrix

In this case, the data are represented as a numeric matrix with one row per sample and one column per state.
The rows must be named according to the corresponding sample in the tree; likewise, the columns must be named according to the discrete state.
The value of the _i,j_<sup>th</sup> element should be 1 if the _i_<sup>th</sup> sample _could be_ state _j_, and 0 otherwise.
I provide an example of a dataset with this format in the package data:
```{r}
# path to character data file
data_file <- system.file("testdata", "sampled_ancestor_data.Rda",  
                         package = "tensoRphylo")

# read the character data
data <- readRDS(data_file)

data
```

## Reading character data

### `NEXUS` format

`tensoRphylo` provides some basic functionality for reading in character data in the desired format.
The function `tensoRphylo::readNexusData` reads a `NEXUS` file with a _single_ discrete character (`nchar = 1`).
This function correctly interprets missing and ambiguous data, and determines the number of states and the state labels from the `NEXUS` file.
Here we read in the sampled ancestor data we saw above ([link to raw file](https://github.com/mikeryanmay/tensoRphylo/blob/main/inst/testdata/sampled_ancestor_data.nex)):
```{r}
# path to character data file
data_file <- system.file("testdata", "sampled_ancestor_data.nex",  
                         package = "tensoRphylo")

# read the character data
data <- readNexusData(data_file)

data
```

### Delimited text format

The function `tensoRphylo::readDelimitedData` reads a comma- or tab-delimited file with one row per sample, the first column containing the sample name, and the second column containing the character data.
The columns _should not_ have headers.
The character data can including missing data (`?` or `-`) or ambiguous data (a `/`-delimited string of compatible states).
The states should be encoded as numeric values starting with 0.
This format does not natively contain information about the state space, so we must provide the number of states to the reader.
Again, we read in the sampled ancestor data we saw above ([link to raw file](https://github.com/mikeryanmay/tensoRphylo/blob/main/inst/testdata/sampled_ancestor_data.csv)).
```{r}
# path to character data file
data_file <- system.file("testdata", "sampled_ancestor_data.csv",  
                         package = "tensoRphylo")

# read the character data
data <- readDelimitedData(data_file, delim = ",", nstates = 2)

data
```

## Conformity

`tensoRphylo` does some basic checks to make sure the tree and data have the same samples.
If you provide a data object with samples that do not occur in the tree, `tensoRphylo` will simply ignore those samples.
Likewise, if you provide a tree with samples that do not occur in the data, `tensoRphylo` will simply add missing data for those samples to the provided dataset.
